Examining background gene lists in clusterProfiler - scRNA-seq example

### 2023-09-06 analysis/example_meth.Rmd
analysis for fig1

---
title: "Examining background gene lists in clusterProfiler - proteomics example"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/enrichment_recipe

@@

suppressPackageStartupMessages({
  library("DESeq2")
  library("kableExtra")
  library("clusterProfiler")
  library("fgsea")
  library("eulerr")
  library("gplots")
})

@@

For this report I will be using proteomics data from Stukalov et al (https://pubmed.ncbi.nlm.nih.gov/33845483/),
who profiled the effect of SARS-COV2 infection on A549 cells.

Here we used the 24hr profile, as there were relatively more changes found.

Probable contaminants were removed, as well as viral proteins.
This left 5825 proteins, including 165 downregulated and 44 upregulated.

The data were obtained from Supplementary Table 5 and saved to the /data folder.

@@

def <- read.table("../data/sarscov2_prot.tsv",sep="\t",header=TRUE)

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 |

@@

Get the gene sets loaded in R.
These are KEGG gene sets

Now filter the gene names into three lists, up-regulated, down-regulated and
background.
Background is simply all genes that were detected.

```{r,subset1}

defup <- unique(subset(def,change.SARS_CoV2.24h_vs_mock.24h=="+")$gene_name)

defdn <- unique(subset(def,change.SARS_CoV2.24h_vs_mock.24h=="-")$gene_name)

bg <- unique(def$gene_name)

message(paste("number of proteins in background:",length(bg)))

Enrichment firstly with upregulated genes

@@

ora_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(ora_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

Now repeat with the downregulated genes

@@

ora_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(ora_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

Now filter the gene names into three lists, up-regulated, down-regulated and
background.
Background is simply all genes that were detected.

@@

Enrichment firstly with upregulated genes

@@

orafix_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(orafix_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

Now with the downregulated genes

@@

orafix_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(orafix_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downregulated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

par(mar=c(10,10,10,10))
par(mfrow=c(2,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated KEGG pathways")
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated KEGG pathways")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")

png("prot_kegg_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 |

@@

head(ora_up) %>%
  kbl(row.names = FALSE, caption="Top upregulated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | Stukalov et al 2021: PMID 33845483 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

par(mar=c(10,10,10,10))
par(mfrow=c(2,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="upregulated Reactomes")
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="downregulated Reactomes")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")

png("prot_reactome_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

>>>>

---
title: "Examining background gene lists in clusterProfiler - methylation example"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/background

@@

suppressPackageStartupMessages({
  library("DESeq2")
  library("kableExtra")
  library("clusterProfiler")
  library("fgsea")
  library("eulerr")
  library("gplots")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
})

@@

For this report I will be using Epigenetics data from an epigenome wide study for smokers versus non-smokers.
PubmedID:27651444
Data was downloaded from EWAS Catalog:
http://ewascatalog.org/?study=27651444_smoking_current_vs_never_smoking

@@

x <- read.table("../data/27651444_smoking_current_vs_never_smoking_16938071371108067.tsv",header=TRUE,sep="\t")

Get background gene list.

```{r,bg}

anno = getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
bg <- anno$UCSC_RefGene_Name
bg <- unique(unlist(strsplit(bg,";")))
length(bg)

Get up and down-methylated list.

```{r,dm1}

defup <- subset(x,beta>0)$gene
defup <- defup[defup!="-"]
defup <- unique(unlist(strsplit(defup,";")))
length(defup)

defdn <- subset(x,beta<0)$gene
defdn <- defdn[defdn!="-"]
defdn <- unique(unlist(strsplit(defdn,";")))
length(defdn)

Gene sets.
Get the gene sets loaded in R.
These are KEGG gene sets

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with one or more probes on the 450K array |
| Gene Selection Criteria | Metaanalysis FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | PMID: 27651444 |
| Other parameters | Min gene set size of 10 |

@@

Enrichment firstly with upmethylated genes

@@

ora_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(ora_up) %>%
  kbl(row.names = FALSE, caption="Top up-methylated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

Now repeat with the downmethylated genes

@@

ora_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(ora_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downmethylated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with one or more probes on the 450K array |
| Gene Selection Criteria | Metaanalysis FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | PMID: 276514443 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

@@

Enrichment firstly with upmethylated genes

@@

orafix_up <- as.data.frame(enricher(gene = defup ,
  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(orafix_up) %>%
  kbl(row.names = FALSE, caption="Top upmethylated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

Now with the downmethylated genes

@@

orafix_dn <- as.data.frame(enricher(gene = defdn ,
  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets2,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

head(orafix_dn,n_pw) %>%
  kbl(row.names = FALSE, caption="Top downmethylated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

par(mar=c(10,10,10,10))
par(mfrow=c(2,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="up-methylated KEGG pathways")
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="down-methylated KEGG pathways")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated KEGG pathways")

svg("meth_kegg_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with one or more probes on the 450K array |
| Gene Selection Criteria | Metaanalysis FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | PMID: 27651444 |
| Other parameters | Min gene set size of 10 |

@@

head(ora_up) %>%
  kbl(row.names = FALSE, caption="Top upmethylated pathways in LPS treatment") %>%
  kable_paper("hover", full_width = F)

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with one or more probes on the 450K array |
| Gene Selection Criteria | Metaanalysis FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | Data availability | PMID: 27651444 |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

par(mar=c(10,10,10,10))
par(mfrow=c(2,1))
plot(euler(v1),quantities = list(cex = 2), labels = list(cex = 2),main="up-methylated Reactomes")
plot(euler(v2),quantities = list(cex = 2), labels = list(cex = 2),main="down-methylated Reactomes")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated Reactomes")

svg("meth_reactome_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-methylated Reactomes")
dev.off()

### 2023-09-06 analysis/example_scrnaseq2.Rmd
analysis for fig1

Source: https://github.com/markziemann/enrichment_recipe

@@

load("muscat_vignette.RData")

ls()

@@

ora_res <- as.data.frame(enricher(gene = fg ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

ora_res <- as.data.frame(enricher(gene = fg ,
  universe = bg,  maxGSSize = 5000, TERM2GENE = genesets_fix,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

Downstream analysis was done according to the Muscat vignette.

Focusing on CD4 T-cells.

@@

lapply(tbl, function(x) {
  nsig <- nrow(subset(x,p_adj.loc<0.05))
  ntot <- nrow(x)
  c(ntot,nsig)
} )

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | Muscat FDR<0.05 CD4 T-cells |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 5 |

@@

def <- lapply(tbl , function(x) {
  bg <- unique(x$gene)
  up <- unique(subset(x,p_adj.loc<0.05&logFC>0)$gene)
  dn <- unique(subset(x,p_adj.loc<0.05&logFC<0)$gene)
  return(list("bg"=bg,"up"=up,"dn"=dn))
} )

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

png("scrna_kegg_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | R/Experiment hub "Kang" or GSE96583. |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

png("scrna_reactome_euler.png")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()

>>>>

Source: https://github.com/markziemann/background

@@

tbl <- readRDS("scDEres.Rds")

@@

ora_res <- as.data.frame(enricher(gene = fg ,
  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

ora_res <- as.data.frame(enricher(gene = fg ,
  universe = bg,  maxGSSize = 500000, TERM2GENE = genesets_fix,
  pAdjustMethod="fdr",  pvalueCutoff = 1, qvalueCutoff = 1  ))

@@

Downstream analysis was done according to the vignette from Seurat.
Link: https://satijalab.org/seurat/archive/v3.1/immune_alignment.html

@@

x <- lapply(tbl, function(x) {
  nsig <- nrow(subset(x,p_val_adj<0.05))
  ntot <- nrow(x)
  c(ntot,nsig)
} )

x

x <- do.call(rbind,x)
colnames(x) <- c("ngenes","nsig")
x <- as.data.frame(x)

x <- subset(x,ngenes>=800 & nsig>=100)

x

x <- rownames(x)

tbl <- tbl[names(tbl) %in% x]

str(tbl)

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | As defined by Seurat |
| Gene Selection Criteria | Seurat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | SeuratData ifnb |
| Other parameters | Min gene set size of 5 |

@@

def <- lapply(tbl , function(x) {
  bg <- unique(rownames(x))
  up <- unique(rownames(subset(x,p_val_adj<0.05 & avg_log2FC>0)))
  dn <- unique(rownames(subset(x,p_val_adj<0.05 & avg_log2FC<0)))
  return(list("bg"=bg,"up"=up,"dn"=dn))
} )

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | KEGG (2023-06-16) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | As defined by Seurat |
| Gene Selection Criteria | Seurat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | SeuratData ifnb |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

svg("scrna2_kegg_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated KEGG pathways")
dev.off()

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list |  As defined by Seurat |
| Gene Selection Criteria | Seurat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | SeuratData ifnb |
| Other parameters | Min gene set size of 10 |

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | As defined by Seurat |
| Gene Selection Criteria | Seurat FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | SeuratData ifnb |
| Other parameters | Min gene set size of 10 + BUG FIX |

@@

svg("scrna2_reactome_euler.svg")
plot(euler(v3),quantities = list(cex = 2), labels = list(cex = 2),main="up- and down-regulated Reactomes")
dev.off()
